---
title: 'R Notebook: colors'
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: true
    number_sections: true 
---

```{r}
library(devtools);
path.github = "https://raw.githubusercontent.com/MonteShaffer/humanVerseWSU/master/";
source_url ( paste0( path.github, "humanVerseWSU/R/functions-colors.R" ) );

```
<http://md5.mshaffer.com/WSU_STATS419/_NOTEBOOKS_/>

# Colors

Here was a color tool I developed in 2007:

<http://colors.mshaffer.com/>

<http://colors.mshaffer.com/?debug=monte>

I tested it in the latest version of Chrome/Edge, still works ...


## Color Palettes in `grDevices`

### Base Colors

There are several color options.  The default colors are stored in the functions `colors()`

These are mostly `x11` colors: <https://en.wikipedia.org/wiki/X11_color_names>

```{r}

library(grDevices);

base.colors = colors();

colors.df = as.data.frame( cbind( (1:length(base.colors)), base.colors) );

colors.df;

  displayColorOptions(base.colors);

```

### `rainbow`

The code below generates 40 contiguous colors using the `?rainbow` function.  These values are hexadecimal RGB format <https://www.w3schools.com/colors/colors_hexadecimal.asp>.
```{r}

colors.rainbow = rainbow(40, s = 0.6, v = 0.75);
  displayColorOptions(colors.rainbow, ncol=4);


```



#### Transparency: `rainbow`

The true RGB color has another element appended, "alpha transparency" also in hexadecimal form, providing a RGBa color <https://www.w3schools.com/Css/css_colors_rgb.asp>.

```{r}

colors.rainbow = rainbow(40, s = 0.6, v = 0.75,
                      alpha=0.5);
  displayColorOptions(colors.rainbow, ncol=4);


```



### `heat.colors`

The code below generates 40 contiguous colors using the `?heat.colors` function.  

```{r}

colors.heat.colors = heat.colors(40);
  displayColorOptions(colors.heat.colors, ncol=4);


```

#### Transparency: `heat.colors`

The code below generates 40 contiguous colors using the `?heat.colors` function.  

```{r}

colors.heat.colors = heat.colors(40, alpha=0.5);
  displayColorOptions(colors.heat.colors, ncol=4);


```



### `terrain.colors`

The code below generates 40 contiguous colors using the `?terrain.colors` function.  

```{r}

colors.terrain.colors = terrain.colors(40);
  displayColorOptions(colors.terrain.colors, ncol=4);


```


#### Transparency: `terrain.colors`

The code below generates 40 contiguous colors using the `?terrain.colors` function.  

```{r}

colors.terrain.colors = terrain.colors(40, alpha=0.5);
  displayColorOptions(colors.terrain.colors, ncol=4);


```



### `topo.colors`

The code below generates 40 contiguous colors using the `?topo.colors` function.  

```{r}

colors.topo.colors = topo.colors(40);
  displayColorOptions(colors.topo.colors, ncol=4);


```


#### Transparency: `topo.colors`

The code below generates 40 contiguous colors using the `?topo.colors` function.  

```{r}

colors.topo.colors = topo.colors(40, alpha=0.5);
  displayColorOptions(colors.topo.colors, ncol=4);


```


### `cm.colors`

The code below generates 40 contiguous colors using the `?cm.colors` function.  

```{r}

colors.cm.colors = cm.colors(40);
  displayColorOptions(colors.cm.colors, ncol=4);


```


#### Transparency: `cm.colors`

The code below generates 40 contiguous colors using the `?cm.colors` function.  

```{r}

colors.cm.colors = cm.colors(40, alpha=0.5);
  displayColorOptions(colors.cm.colors, ncol=4);


```

























### Color Palettes using `hcl.colors`


```{r}

hcl.palette.list = hcl.pals();

as.data.frame( hcl.palette.list );

for(i in 1:length(hcl.palette.list))
  {
  hcl = hcl.palette.list[i];
  colors.hcl.colors = hcl.colors(40, palette=hcl);
  
  print(paste0("Palette: ", hcl));
    displayColorOptions(colors.hcl.colors, ncol=4);
  print(paste0("   ---> alpha: ", 0.5));
  colors.hcl.colors = hcl.colors(40, palette=hcl,
                                        alpha=0.5);
    displayColorOptions(colors.hcl.colors, ncol=4);
  }

```



### Build a color gradient using `colorRampPalette`

You can interpolate a color gradient as well.

```{r}

colvec=c("red","royalblue");
my.colors = colorsInGradient(40, colvec, FALSE);
  displayColorOptions(my.colors, ncol=4);


colvec=c("red","royalblue","seagreen");
my.colors = colorsInGradient(40, colvec, FALSE);
  displayColorOptions(my.colors, ncol=4);

```

### Helper functions

I wrote a few helper functions to manipulate and store colors.

- `displayColorOptions` to display these rectangular summaries of colors.  If rendered as HTML, it would be more useful, as you could copy/paste the code.  I will work on that option.

- `colorsInGradient` is a wrapper for the odd-structured function `colorRampPalette`

- There is a built-in `col2rgb` function that was missing its reciprocal or inverse, so I wrote it as `rgb2col` 

`col2rgb("red");`  and `rgb2col( col2rgb("red") );`  ... useful for grabbing the HEX color scheme from the named scheme

- A lookup tool on the many (600+ colors).  `indexColors` builds the index and `searchColor` can do a lookup based on name (which is unique) or based on hex (which is not necessarily unique, so may return multiple values).  I may use a multi-dimensional distance metric to compute a "fuzzyColorSearch" calling it `findNearestColor`

# Conclusion

Online, you can read about all these special/cool palettes developed by others libraries (e.g.,  "RColorBrewer" or "wesanderson").  There is nothing new or novel in these libraries.  RGB and specifically RGBa are how colors are built in R (and in most user interfaces).  You can do some research, and build your own palette using the information contained herein.  The "pretty colors" are often a function of an "alpha" level of RGBa.  And it is related to the psychological perceptions we have as to why the color is "pretty".

```{r}

# build a custom palette ...

# https://brand.wsu.edu/visual/colors/

colors.list = list();
colors.list[["monte_shaffer@wsu.edu"]] = list();

colors.list[["monte_shaffer@wsu.edu"]][["crimson"]] = "#981e32";
colors.list[["monte_shaffer@wsu.edu"]][["crimson-accent"]] = "#c60c30";
colors.list[["monte_shaffer@wsu.edu"]][["gray"]] = "#5e6a71";
colors.list[["monte_shaffer@wsu.edu"]][["gray-darker1"]] = "#464e54";
colors.list[["monte_shaffer@wsu.edu"]][["gray-darker2"]] = "#2a3033";
colors.list[["monte_shaffer@wsu.edu"]][["gray-black"]] = "#000000";
colors.list[["monte_shaffer@wsu.edu"]][["gray-lighter1"]] = "#8d959a";
colors.list[["monte_shaffer@wsu.edu"]][["gray-lighter2"]] = "#b5babe";
colors.list[["monte_shaffer@wsu.edu"]][["gray-lighter3"]] = "#d7dadb";
colors.list[["monte_shaffer@wsu.edu"]][["gray-lighter4"]] = "#eff0f1";
colors.list[["monte_shaffer@wsu.edu"]][["gray-white"]] = "#FFFFFF";
colors.list[["monte_shaffer@wsu.edu"]][["gray-lighter1"]] = "#8d959a";


colors.list[["monte_shaffer@wsu.edu"]][["orange"]] = "#b67233";
colors.list[["monte_shaffer@wsu.edu"]][["orange-accent"]] = "#f6861f";

colors.list[["monte_shaffer@wsu.edu"]][["green"]] = "#8f7e35";
colors.list[["monte_shaffer@wsu.edu"]][["green-accent"]] = "#ada400";

colors.list[["monte_shaffer@wsu.edu"]][["blue"]] = "#4f868e";
colors.list[["monte_shaffer@wsu.edu"]][["blue-accent"]] = "#00a5bd";

colors.list[["monte_shaffer@wsu.edu"]][["yellow"]] = "#c69214";
colors.list[["monte_shaffer@wsu.edu"]][["yellow-accent"]] = "#ffb81c";

# to practice, build out your own color scheme ... 

mycolors = unlist(colors.list$`monte_shaffer@wsu.edu`);

displayColorOptions(mycolors, showHEX=TRUE, ncol=2);


```

When you play with RGB addressable lights in electronics, you can get two types, RGB and RGBw.  The ability to create mixture of light by having 3 or 4 small LEDs light up is the basic function of how light emission works.  Here is a short video clip <https://www.youtube.com/watch?v=DBRJ0KbI1Hk> of an application, which was programmed in embedded C as part of a research grant to develop automated technologies to engage the mind and body of persons with Parkinson's disease <https://www.nsf.gov/awardsearch/showAward?AWD_ID=1819997>.  [We also had a nice punch-detection algorithm that would use the statistic called MAD to perform real-time signal processing.  I will likely document some of that algorithm in a "mastery" notebook.]

Obviously, the color schema presented herein is a function of light emission (which is how computers operate) and the aesthetics would be different for a paint-schema for color, as that is a function of how light reflects off a tangible surface.