---
title: 'R Notebook: using Python in R'
output:
  html_document:
    df_print: paged 
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: true
    number_sections: true 
---

```{r}

library(devtools);

library(humanVerseWSU);

path.github = "https://raw.githubusercontent.com/MonteShaffer/humanVerseWSU/master/";

include.me = paste0(path.github, "misc/functions-nlp.R");
# source_url( include.me );
include.me = paste0(path.github, "humanVerseWSU/R/functions-encryption.R");
source_url( include.me );



path.to.nascent = "C:/Users/Alexander Nevsky/Dropbox/WSU-419/Fall 2020/__student_access__/unit_02_confirmatory_data_analysis/nascent/";

folder.nlp = "nlp/";
path.to.nlp = paste0(path.to.nascent, folder.nlp);


```
# Using Python with R

In this notebook, we will briefly discuss how to setup your workspace to use Python inside of R.

The instructions are for a Windows 10 platform, but the concepts and themes can be applied to a MacOS with some a bit of web searching and self-directed problem solving.

## Download and Install Python

<https://www.python.org/downloads/windows/>.  

- I would install it to `C:\python3`
- I would add to "Settings" -> Environmental Variables -> Path
- Now it is "python" not "python2" or "python3" from a shell.

## Download and Install PIP

The package manager for Python is called "pip" ... there are other alternatives, but "pip" has been around the longest.

- Download <https://bootstrap.pypa.io/get-pip.py> to `C:\python3`
- Run from CMD `C:\python3`: `python get-pip.py`
- Add "pip" to "Settings" -> Environmental Variables -> Path `C:\python3\Scripts`
- Close and Open CMD.  Now type `pip install --upgrade pip setuptools wheel` to verify that things are up to date and working.

**You may need to restart RStudio**

## Download and install packages

- Open a CMD prompt
- Type `pip install -U scikit-learn` to install `scikit-learn` (see: <https://scikit-learn.org/stable/install.html>) 
- Type `pip install -U nltk` to install `nltk` (see: <https://www.nltk.org/>) 
- Type `pip install -U numpy` to install `numpy` (see: <https://numpy.org/>)

**Windows C++ compiler**

If you get an error, you may need to download an extra compiler.  I would recommend minimizing the download unless you use Microsoft Tools frequently.

Visual C++ 14.0 tools! <https://stackoverflow.com/questions/44951456/pip-error-microsoft-visual-c-14-0-is-required> which has this link <http://go.microsoft.com/fwlink/?LinkId=691126&fixForIE=.exe.>

## Enable Python in Your Notebook

```{r}
library(reticulate);
use_python("C:/python3");
```

## Hello World in Python

```{python}

print("hello world");

def pownum(base, pow):
  return base ** pow

print(pownum(9,3));

```
## Python IDE:  WingIDE

This seems to be a nice IDE.  The PRO version is free for education purposes.  I have requested a key-code for the course and will share once I receive it.

<https://wingware.com/> 

