---
title: 'R Notebook: bits and such)'
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: true
    number_sections: true 
---

```{r}

library(devtools);

library(humanVerseWSU);

path.github = "https://raw.githubusercontent.com/MonteShaffer/humanVerseWSU/master/";


include.me = paste0(path.github, "humanVerseWSU/R/functions-encryption.R");
source_url( include.me );

include.me = paste0(path.github, "humanVerseWSU/R/functions-number.R");
source_url( include.me );

```

# The 'md5' algorithm

Saturday, I had a research meeting cancelled, so I said "hey, let me spend a couple of hours reviewing the md5."

The "best laid schemes of mice and men often go astray."  This was a John Steinbeck quote from a poem by Robert Burns.  It just means "things don't often go as planned."

So I ran into a "bit-wise" issue, and I thought I would share that.


## To understand the "md5"

The website <http://md5.mshaffer.com/> allows you to enter a string of text and it converts it to a 32-digit hexadecimal string.  Until recently, there was not thought to be any collisions.


- MD5: The fastest and shortest generated hash (16 bytes). The probability of just two hashes accidentally colliding is approximately: 1.47*10-29.

- SHA1: Is generally 20% slower than md5, the generated hash is a bit longer than MD5 (20 bytes). The probability of just two hashes accidentally colliding is approximately: 1*10-45

- SHA256: The slowest, usually 60% slower than md5, and the longest generated hash (32 bytes). The probability of just two hashes accidentally colliding is approximately: 4.3*10-60.

Source: <https://www.avira.com/en/blog/md5-the-broken-algorithm>

## Collision "pandemonian"

You can read about hacking "md5" to introduce collisions on "file" md5sums, which is a function in tools.  `tools::md5sum`

An informed discussion on the "collision hysteria":

- <https://stackoverflow.com/questions/933497/create-your-own-md5-collisions> **Can't be done with strings**

- <https://security.stackexchange.com/questions/37454/why-are-md5-collisions-dangerous> **Is dangerous if you use them for passwords**

I laugh because storing and md5(password) in a database was a "secure" method about 15 years ago.  Now, we store salt and pepper.  And by we, I mean people that know what they are doing, not Facebook <https://www.npr.org/2019/03/21/705588364/facebook-stored-millions-of-user-passwords-in-plain-readable-text>.

For a unique representation of a string, `md5` is the way to go.

## RFC on md5 

The open-source community I grew up in had this concept of RFC:  request for comments.  The community together built the key features of important software.  Here is the April 1992 RFC-1321 for the md5: <https://tools.ietf.org/html/rfc1321>.

- Page 12 shows the basic transformation
- Page 14 shows the basic string output from the digest
- Page 20 shows some test examples

```{r}
tests = c("", "a", "abc", "message digest", "abcdefghijklmnopqrstuvwxyz", "ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz0123456789", "123456789012345678901234567890123456789012345678901234567890123456
78901234567890");

# monte adds
tests = c(tests, "monte", "alex", "The quick brown fox jumps over the lazy dog", "The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog :: The quick brown fox jumps over the lazy dog")

```

## Javascript c-code

At some point, I stumbled across this code (circa 2005): <http://md5.mshaffer.com/md5.js>.  It was compact with subfunctions.  Then Jquery came out with digest algorithms packaged together and much faster, so I abandonened it.

Anyway, I thought, "hey I got two hours, let me code this out in R."  I ran into some bit issues, but it can be found as the function `.md5` in the `functions-encryption.R` (not yet compiled, so you have to view the source).  The issues were related to "signed" bit operations.

If you compare the code to the RFC, you will notice it is very similar.

<pre>
// http://block111.servehttp.com/js/md5.js
function md5(s){
	function Z(n,c){return(n<<c)|(n>>>(32-c))}
	function Y(q,a,b,x,s,t){return X(Z(X(X(a,q),X(x,t)),s),b)}
	function A(a,b,c,d,x,s,t){return Y((b&c)|((~b)&d),a,b,x,s,t)}
	function B(a,b,c,d,x,s,t){return Y((b&d)|(c&(~d)),a,b,x,s,t)}
	function C(a,b,c,d,x,s,t){return Y(b^c^d,a,b,x,s,t)}
	function D(a,b,c,d,x,s,t){return Y(c^(b|(~d)),a,b,x,s,t)}
	function X(x,y){var l=(x&0xFFFF)+(y&0xFFFF),m=(x>>16)+(y>>16)+(l>>16);return(m<<16)|(l&0xFFFF)}
	
	var w=s.length*8,i,hx="0123456789abcdef",L=(((w+64)>>>9)<<4)+15,x=Array(L+15);
	for(i=0;i<x.length;++i)x[i]=0;
	for(i=0;i<w;i+=8)x[i>>5]|=(s.charCodeAt(i/8)&255)<<(i%32);
	x[w>>5]|=0x80<<((w)%32);
	x[L-1]=w;
	
	var a=1732584193,b=-271733879,c=-1732584194,d=271733878;
	for(i=0;i<L;i+=16){
		var oa=a,ob=b,oc=c,od=d;
		a=A(a,b,c,d,x[i],7,-680876936);d=A(d,a,b,c,x[i+1],12,-389564586);c=A(c,d,a,b,x[i+2],17,606105819);b=A(b,c,d,a,x[i+3],22,-1044525330);
		a=A(a,b,c,d,x[i+4],7,-176418897);d=A(d,a,b,c,x[i+5],12,1200080426);c=A(c,d,a,b,x[i+6],17,-1473231341);b=A(b,c,d,a,x[i+7],22,-45705983);
		a=A(a,b,c,d,x[i+8],7,1770035416);d=A(d,a,b,c,x[i+9],12,-1958414417);c=A(c,d,a,b,x[i+10],17,-42063);b=A(b,c,d,a,x[i+11],22,-1990404162);
		a=A(a,b,c,d,x[i+12],7,1804603682);d=A(d,a,b,c,x[i+13],12,-40341101);c=A(c,d,a,b,x[i+14],17,-1502002290);b=A(b,c,d,a,x[i+15],22,1236535329);
		a=B(a,b,c,d,x[i+1],5,-165796510);d=B(d,a,b,c,x[i+6],9,-1069501632);c=B(c,d,a,b,x[i+11],14,643717713);b=B(b,c,d,a,x[i],20,-373897302);
		a=B(a,b,c,d,x[i+5],5,-701558691);d=B(d,a,b,c,x[i+10],9,38016083);c=B(c,d,a,b,x[i+15],14,-660478335);b=B(b,c,d,a,x[i+4],20,-405537848);
		a=B(a,b,c,d,x[i+9],5,568446438);d=B(d,a,b,c,x[i+14],9,-1019803690);c=B(c,d,a,b,x[i+3],14,-187363961);b=B(b,c,d,a,x[i+8],20,1163531501);
		a=B(a,b,c,d,x[i+13],5,-1444681467);d=B(d,a,b,c,x[i+2],9,-51403784);c=B(c,d,a,b,x[i+7],14,1735328473);b=B(b,c,d,a,x[i+12],20,-1926607734);
		a=C(a,b,c,d,x[i+5],4,-378558);d=C(d,a,b,c,x[i+8],11,-2022574463);c=C(c,d,a,b,x[i+11],16,1839030562);b=C(b,c,d,a,x[i+14],23,-35309556);
		a=C(a,b,c,d,x[i+1],4,-1530992060);d=C(d,a,b,c,x[i+4],11,1272893353);c=C(c,d,a,b,x[i+7],16,-155497632);b=C(b,c,d,a,x[i+10],23,-1094730640);
		a=C(a,b,c,d,x[i+13],4,681279174);d=C(d,a,b,c,x[i],11,-358537222);c=C(c,d,a,b,x[i+3],16,-722521979);b=C(b,c,d,a,x[i+6],23,76029189);
		a=C(a,b,c,d,x[i+9],4,-640364487);d=C(d,a,b,c,x[i+12],11,-421815835);c=C(c,d,a,b,x[i+15],16,530742520);b=C(b,c,d,a,x[i+2],23,-995338651);
		a=D(a,b,c,d,x[i],6,-198630844);d=D(d,a,b,c,x[i+7],10,1126891415);c=D(c,d,a,b,x[i+14],15,-1416354905);b=D(b,c,d,a,x[i+5],21,-57434055);
		a=D(a,b,c,d,x[i+12],6,1700485571);d=D(d,a,b,c,x[i+3],10,-1894986606);c=D(c,d,a,b,x[i+10],15,-1051523);b=D(b,c,d,a,x[i+1],21,-2054922799);
		a=D(a,b,c,d,x[i+8],6,1873313359);d=D(d,a,b,c,x[i+15],10,-30611744);c=D(c,d,a,b,x[i+6],15,-1560198380);b=D(b,c,d,a,x[i+13],21,1309151649);
		a=D(a,b,c,d,x[i+4],6,-145523070);d=D(d,a,b,c,x[i+11],10,-1120210379);c=D(c,d,a,b,x[i+2],15,718787259);b=D(b,c,d,a,x[i+9],21,-343485551);
		
		a=X(a,oa);
		b=X(b,ob);
		c=X(c,oc);
		d=X(d,od);
	}
	b=[a,b,c,d];
	console.log(b);
	s="";
	for(i=0;i<16;i++)s+=hx.charAt((b[i>>2]>>((i%4)*8+4))&0xF)+hx.charAt((b[i>>2]>>((i%4)*8))&0xF);
	return s
}
</pre>

## Help on bit operations

So I was in the process of "porting" the code over, and ran into some issues.  That opens a "can of worms" on how to jump into the functions and debug.  

<https://stat.ethz.ch/R-manual/R-devel/library/base/html/bitwise.html>

**Shifting is done assuming the values represent unsigned integers.**

Above is the issue.  R is older than RFC-1321.  And its bit-wise operators are a bit ancient.  So I asked a question:

<https://stackoverflow.com/questions/64839024/using-r-how-to-do-bitwise-shifting-for-signed-negative-integers/>

And I learned a few things on my extra-hour adventure:

```{r}
# if you have Rtools and Rcpp in place:

# Run these two lines and the "c++" form of the function is at your disposal in R.

####### Rcpp::cppFunction("long long ShiftR(long long a, int b) { return a >> b;}");
####### Rcpp::cppFunction("long long ShiftL(long long a, int b) { return a << b;}");

```

There are also issues of overflow on "shiftL" and I believe I have hacked together a solution.  Not fully tested, but seems to perform `.md5` okay.

```{r}

bitShiftL = function(x, bits, unsigned=FALSE)
  { 
  # look it up, currently a hack
  }

bitShiftR = function(x, bits, unsigned=FALSE)
  {
  # as it gets to big, goes to "0" for positive, "-1" for negative: the signed bit
  if(!is.negative(x) | unsigned) { return( bitwShiftR(x,bits) ); }
  -bitwShiftR(-x,bits) - 1;
  }

bitOr = function(a, b)
  {
  # look it up, currently a hack
  }

is.negative = function(x, tol = sqrt(.Machine$double.eps))
  {
  x < ( -1 * tol );
  }

include.me = paste0(path.github, "humanVerseWSU/R/functions-encryption.R");
source_url( include.me );

include.me = paste0(path.github, "humanVerseWSU/R/functions-number.R");
source_url( include.me );

```

Like the `isClose` function, `is.negative` does not assume much about the actual value.  Some of these ideas came from help on `?is.integer` which is not `is.whole.number` so I wrote that function as well.

## Benchmarking with microben

```{r}
# Note: I loop over a single string at a time because digest does not have vector capabilities, it just returns md5 of first.
    #  humanVerseWSU::md5(s) is a wrapper for digest (vectors not allowed)
    #  nmd5(s) is a wrapper for openssl (vectors allowed)

## this is slow because my function is very slow...

library(microbenchmark);
library(openssl);
library(digest);

s = "The quick brown fox jumps over the lazy dog";
res = list();
i = 1;
#for(s in tests)
  {
  print(s);
  mb = microbenchmark(                            .md5(s),
                                          openssl::md5(s),
                              # wrapper for openssl (vector)
                                                  nmd5(s), 
           digest::digest(s, algo="md5", serialize=FALSE),
                              # wrapper for digest (non-vector)
                                    humanVerseWSU::md5(s),  
                times=25 );
  
  res[[i]] = mb;
  i = 1 + i;
  }

 # rownames(mb) = mb$expr;
summary(mb);

```


We see that digest is "fast" yet it does not allow "vector" formats.  My "hacked" approach works, but is very slow.  Recall:  I also demonstrated "old-school" sorting techniques that are also very slow.  As computing gets more complicated, it is a challenge to delve in and understand the features of each.

### digest

<https://cran.r-project.org/web/packages/digest/index.html>

Digest is very fast, but not "variadic" --- I have to pass in one value at a time.  I can wrap it in my own function "md5" or use an "apply" macro, but that is really not a pleasant design.  I read through the code base a bit, and it is not easy to identify where the algorithms for a given hash is found

### openssl

This library is a wrapper for the c-based library that has been in existence for about 20 years.  I am well aware of the various encryption/decryption protocols of the core engine and have used them on several projects.  

### Final answer

For me, I will migrate away from digest to openssl for several reasons:

- digest doesn't have vector form (not variadic)
- openssl is a wrapper for a well-established c-platform
- openssl participates frequently in RFC
- openssl had to deal with serious hacks (see <https://en.wikipedia.org/wiki/Heartbleed>).  Very likely, NSA was exploiting this vulnerability for some time (see <https://en.wikipedia.org/wiki/Heartbleed#Possible_prior_knowledge_and_exploitation>).  If you are fighting head-to-head with NSA and winning most of the time, it is a good library.



